A Constraint Solving Approach to Parikh Images of Regular Languages

### 2023-06-24 calculus.tex
An acceptable (but hardly great) example
This also adds propagation to 2.1, which I had forgotten.

Punching out from today after 5h 30 min.

\subsection{An Example}
Here we will return to $A'$ from \cref{sec:intuition} and performs the steps of
\cref{sec:intuition:algebra,sec:intuition:prop,sec:intuition:split} with the
formal calculus we just established.

Starting with~$A'$, $\Map$ extracts the increments of a transition (see
\cref{fig:a_1}), $h(\FromLabelTo{q_1}{c / i}{q_2}) = \Vec{v}$ for the increment
vector $\Vec{v}$, concretely $h(\FromLabelTo{I_A}{\Sigma / \left[ l_{A+}, i_+
\right]}{I_A}) = \left[ l_{A+}, i_+ \right]$. We use the same compact notation
here as in \cref{sec:intuition} to represent what is essentially a sparse vector
of $1$ and $0$ coefficients. Initially, we let $\Filter$ map to fresh variables
to obtain:
\begin{equation*}
    \begin{aligned}
      \Filter(\FromLabelTo{D}{d/\begin{bmatrix}l_{A+}, n_+\end{bmatrix}}{F_A}) & = \TransitionVar_1    & 
                \Filter(\FromLabelTo{F_A}{\Sigma/\begin{bmatrix} l_{A+} \end{bmatrix}}{F_A} ) & = \TransitionVar_2 \\
      \Filter(\FromLabelTo{I_A}{c/\begin{bmatrix}l_{A+}, n_+\end{bmatrix}}{F_A} ) & =  \TransitionVar_3 & 
                \Filter(\FromLabelTo{I_A}{d/\begin{bmatrix}l_{A+}, n_+\end{bmatrix}}{D}) & = \TransitionVar_4 \\
      \Filter(\FromLabelTo{I_A}{\Sigma/\begin{bmatrix} l_{A+} i_+\end{bmatrix}}{I_A}) & = \TransitionVar_5 & 
      \FlowEq(A', \Filter) & = \begin{aligned}
        1 + \cancel{\TransitionVar_5} &= \cancel{\TransitionVar_5} + \TransitionVar_3 \\
        \TransitionVar_4 &= \TransitionVar_1 \\
        \TransitionVar_3 + \TransitionVar_1 + \cancel{\TransitionVar_2} &\geq \cancel{\TransitionVar_2}  \\
      \end{aligned} \\
    \end{aligned}
\end{equation*}

% RUN EXPAND GET:
$i = \TransitionVar_5, l_A = \sum_{k = 1}^{k = 5} \TransitionVar_k, n = \TransitionVar_4 + \TransitionVar_1$

  If we apply reasoning by equivalence as in \cref{sec:a_1}, we end up with the
  following definitions, corresponding to \cref{fig:a_1} at the end of
  \cref{sec:a_1}:
  \begin{equation*}
    \begin{aligned} % Post-propagation
      \Filter(\FromLabelTo{D}{d/\begin{bmatrix}l_{A+}, n_+\end{bmatrix}}{F_A}) = \Filter(\FromLabelTo{I_A}{d/\begin{bmatrix}l_{A+}, n_+\end{bmatrix}}{D}) & = \TransitionVar_1\\
      \Filter(\FromLabelTo{F_A}{\Sigma/\begin{bmatrix} l_{A+} \end{bmatrix}}{F_A} ) & = l_A  - i - n\\
      \Filter(\FromLabelTo{I_A}{c/\begin{bmatrix}l_{A+}, n_+\end{bmatrix}}{F_A} ) & =  1 - \TransitionVar_1\\
      \Filter(\FromLabelTo{I_A}{\Sigma/\begin{bmatrix} l_{A+} i_+\end{bmatrix}}{I_A}) & = i\\
    \end{aligned}
  \end{equation*}

\begin{figure}
  \centering
\begin{prooftree}
  \hypo[]{
{    \begin{array}{lll}
      1 - \TransitionVar_6 = 0 & \land \TransitionVar_6 = 1 & \land  \TransitionVar_4 + 1 > 0 \\
      \land \TransitionVar_3 = 0 &\land  2 + \TransitionVar_4 > 0 & \land k = 1 + \TransitionVar_4 
    \end{array}}
  }
  \infer1[\Subsume{}]{
{    \begin{array}{lll}
      \Connected(\AcaOrBc{}, \Filter) & \land 1 - \TransitionVar_6 = 0 \\
      \land \TransitionVar_6 = 1 & \land \TransitionVar_4 + 1 > 0 \land \\
      \land \TransitionVar_3 = 0 & \land 2 + \TransitionVar_4 > 0 \land \\
      k = 1 + \TransitionVar_4 
    \end{array}}
  }
  \infer1[\Propagate]{
    \begin{aligned}
      \Connected(\AcaOrBc{}, \Filter) \land \\
      1 - \TransitionVar_6 = 0 \land \\
      \TransitionVar_6 = 1 \land \\
      \TransitionVar_4 + 1 > 0 \land \\
      \TransitionVar_3 \geq 0 \land \\
      2 + \TransitionVar_4 \geq 0 \land \\
      2k = 2 + 2\TransitionVar_4 + \TransitionVar_3
    \end{aligned}
  }
  \infer1[\EquationReasoning{}]{
    \begin{aligned}
      \Connected(\AcaOrBc{}, \Filter) \land \\
      1 - \TransitionVar_6 = 0 \land \\
      2k = 1 + 2\TransitionVar_4 + \TransitionVar_3 + \TransitionVar_6
    \end{aligned}
  }
  % BRANCH:  = 0'
  \hypo{
    {\begin{array}{lll}
      \TransitionVar_3 > 0 & \land 
      \TransitionVar_6 = 0 \\ \land
      2\TransitionVar_6 + \TransitionVar_4 = 0 & \land
      1 - \TransitionVar_6 > 0\\  \land 
      2k = 1 + \TransitionVar_3
    \end{array}}
  }
  \infer1[\Subsume{}]{
    \begin{aligned}
      \Connected(\AcaOrBc{}, \Filter) \land \\
      \TransitionVar_3 > = 0 \land \\
      \TransitionVar_6 = 0 \land \\
      2\TransitionVar_6 + \TransitionVar_4 = 0 \land \\
      1 - \TransitionVar_6 > 0 \land \\
      2k = 1 + \TransitionVar_3
    \end{aligned}
  }
  \infer1[\EquationReasoning{}]{
    \begin{aligned}
      \Connected(\AcaOrBc{}, \Filter) \land \\
      \TransitionVar_6 = 0 \land \\
      1 - \TransitionVar_6 > 0 \land \\
      2\TransitionVar_6 + \TransitionVar_4 = 0\\
      2k = 1 + 2\TransitionVar_4 + \TransitionVar_3
    \end{aligned}
  }
  \infer1[\Propagate{}]{
    \begin{aligned}
      \Connected(\AcaOrBc{}, \Filter) \land \\
      \TransitionVar_6 = 0 \land \\
      1 - \TransitionVar_6 > 0 \land \\
      2k = 1 + 2\TransitionVar_4 + \TransitionVar_3
    \end{aligned}
  }
  \infer1[\EquationReasoning{}]{
    \begin{aligned}
      \Connected(\AcaOrBc{}, \Filter) \land \\
      1 - \TransitionVar_6 > 0 \land \\
      2k = 1 + 2\TransitionVar_4 + \TransitionVar_3 + \TransitionVar_6
    \end{aligned} % BRANCH: > 0
  } % SPLIT
  \infer2[\Split{} $1 - \TransitionVar_6$]{ \Connected(\AcaOrBc{}, \Filter) \land 2k = 1 + 2\TransitionVar_4 + \TransitionVar_3 + \TransitionVar_6 }
  \infer1[\EquationReasoning{}]{
    \Connected(\AcaOrBc{}, \Filter) \land
    2k + 1 =
    \sum\limits_{\Transition \in \Transitions} (\Transition \mapsto 1)(\Label) \cdot \Filter(\Transition)
  }
  \infer1[\Expand{}]{\Image{}_{\AcaOrBc{}, \Transition \mapsto 1}(\Filter, 2k + 1)}
\end{prooftree}
\caption{A derivation for \Calculus{} computing odd lengths in $\AcaOrBc{}$.}\label{fig:derivation:single}
\end{figure}

In \cref{fig:derivation:single}, we see how we start by expanding the predicate
using the (simplified) flow equations. We can also see how interleaving
reasoning on the corresponding linear equations helps the deduction. In both
branches the reasoning is similar: we conclude that when choosing either path at
the starting state we must avoid the other one, perform propagation based on
that fact, use some algebraic reasoning to derive the expected forms of the
upper and lower bounds to propagate the now disconnected transitions, and
finally subsume, removing the $\Connected{}$ predicate when we are unable to use
either of the rules, leaving only a set of linear equations. To obtain a solution from the
proof, one would need to perform standard model generation on the remaining
constraints to obtain values for the sought variable $\MonoidElement$. The full
Parikh image is the disjunction of the constraints left at the two
leaves after quantifier elimination.

Following the values of $\Filter$, we can read out the path corresponding to our solution in the product, and with that a word. In our case, for the right branch we have $\Filter(\FromLabelTo{S}{b}{B}) = 0,
\Filter(\FromLabelTo{A}{c}{A})  > 0,
\Filter(\FromLabelTo{B}{b}{S}) = 0,
\Filter(\FromLabelTo{S}{a}{A}) = 1 ,
\Filter(\FromLabelTo{A}{a}{F}) = 1 ,
\Filter(\FromLabelTo{B}{c}{F}) = 0$. We can read out a word along it: aca, accca, etc, and the corresponding constraint: $\exists k : \TransitionVar_3 = 2k -1 \land \TransitionVar_3 > 0 \land k \geq 0$.

>>>>

\subsection{An Example}
Here we will return to $A'$ from \cref{sec:intuition} and performs the steps of
\cref{sec:intuition:algebra,sec:intuition:prop,sec:intuition:split} with the
formal calculus we just established, but exclude \cref{const:s2-in-b} and
therefore the entire automaton $B$ since we introduce support for products of
automata in \cref{sec:multiple}. Note that this example then becomes
satisfiable, as $s_1 = \Char{dd}, i = 1, s_2 = \Char{adda}, l_A = 4$ is a
satisfying assignment to
\cref{const:more-inside-than-before,const:s1-in-c-dd,const:s1-substring}.

Starting with~$A'$, $\Map$ extracts the increments of a transition (see
\cref{fig:a_1}), $h(\FromLabelTo{q_1}{c / i}{q_2}) = \Vec{v}$ for the increment
vector $\Vec{v}$, concretely $h(\FromLabelTo{I_A}{\Sigma / \left[ l_{A+}, i_+
\right]}{I_A}) = \left[ l_{A+}, i_+ \right]$. We use the same compact notation
here as in \cref{sec:intuition} to represent what is essentially a sparse vector
of $1$ and $0$ coefficients, i.e. = $\left[ l_{A+}, i_+ \right] = \left[1, 1,
0\right]$. Initially, we let $\Filter$ map to fresh variables to obtain:
\begin{equation}
    \begin{aligned}
      \Filter(\FromLabelTo{D}{d/\begin{bmatrix}l_{A+}, n_+\end{bmatrix}}{F_A}) & = \TransitionVar_1    & 
                \Filter(\FromLabelTo{F_A}{\Sigma/\begin{bmatrix} l_{A+} \end{bmatrix}}{F_A} ) & = \TransitionVar_2 &       \FlowEq(A', \Filter) & = \begin{aligned}
                    1 + \cancel{\TransitionVar_5} &= \cancel{\TransitionVar_5} + \TransitionVar_3 + \TransitionVar_1\\
                    \TransitionVar_4 &= \TransitionVar_1 \\
                    \TransitionVar_3 + \TransitionVar_1 + \cancel{\TransitionVar_2} &\geq \cancel{\TransitionVar_2}  \\
                  \end{aligned} \\
      \Filter(\FromLabelTo{I_A}{c/\begin{bmatrix}l_{A+}, n_+\end{bmatrix}}{F_A} ) & =  \TransitionVar_3 & 
                \Filter(\FromLabelTo{I_A}{d/\begin{bmatrix}l_{A+}, n_+\end{bmatrix}}{D}) & = \TransitionVar_4 &\\
      \Filter(\FromLabelTo{I_A}{\Sigma/\begin{bmatrix} l_{A+} i_+\end{bmatrix}}{I_A}) & = \TransitionVar_5 & 
    \end{aligned}
\end{equation}\label{eq:single-filter}

Like in \cref{sec:a_1,sec:intuition:algebra}, we apply reasoning by equivalence
  to end up with fixed values for several of the transition variables
  $\TransitionVar_k$, and eventually we use \cref{const:more-inside-than-before}
  and \cref{const:something-before-and-after} to conclude that the transition
  directly from state $I_{A}$ to $F_{A}$ is incompatible with the constraint.

  Then, following \cref{sec:intuition:subsume}, we remove the $\Connected{}$
   predicate using \Subsume{} since no transition can can become disconnected
   (and in fact never could for $A'$, since all looping transitions are attached
   to the initial and single accepting state and therefore always must be
   reachable from any accepting path). After this, the rest of the reasoning can
   be continued without any part of our calculus, since all values of the
   existentially quantified transition variables
   $\TransitionVar_1,\ldots,\TransitionVar_5$ compatible with
   \cref{const:more-inside-than-before,const:s1-in-c-dd,const:s1-substring,const:something-before-and-after}
   are represented by the flow equations introduced by $\Expand{}$. The
   derivation is shown in \cref{fig:derivation:single}.

\begin{figure}
  \centering
\begin{prooftree}
  \hypo[]{{\text{Regular solving finishes the proof}}}
  \infer1[]{
    \begin{aligned}
    \TransitionVar_1 = 1 \land 
    i = \TransitionVar_5 \land 
    n = 2 \land \\
    2 \geq i \land l_A - i - 2 > 0 
    \land \TransitionVar_4 = 1 
    \land \TransitionVar_2 = l_A - i - 2 
    \land \TransitionVar_3 = 0 \\
    \end{aligned}}
  \infer1[\Subsume: \cref{sec:intuition:subsume}]{
    \begin{aligned}
    \Connected(A', \Filter) \land 
    \TransitionVar_1 = 1 \land 
    i = \TransitionVar_5 \land 
    n = 2 \land 
    2 \geq i \land l_A - i - 2 > 0  \\
    \land \TransitionVar_4 = 1 
    \land \TransitionVar_2 = l_A - i - 2 
    \land \TransitionVar_3 = 0 \\
    \end{aligned}
  }
  \infer1[\EquationReasoning{}: $\TransitionVar_1 = 1$ \cref{sec:a_1}]{
    \begin{aligned}
    \Connected(A', \Filter) \land \\
    i = \TransitionVar_5 \land 
    n = 2\TransitionVar_1 \land n \geq i \land l_A - i - n > 0 
    \land \TransitionVar_4 = \TransitionVar_1 \\
    \land \TransitionVar_2 = l_A - i - n 
    \land \TransitionVar_3 = 1 - \TransitionVar_1 
    \end{aligned}
  }
  \infer1[\EquationReasoning{}: $=$-elimination]{
    \begin{aligned}
    \Connected(A', \Filter) \land \\
    i = \TransitionVar_5 \land 
    n = \TransitionVar_4 + \TransitionVar_1 \land n \geq i \land l_A - i - n > 0 \\
    \land \TransitionVar_4 = \TransitionVar_1 
    \land \TransitionVar_2 = l_A - i - n 
    \land \TransitionVar_3 = 1 - \TransitionVar_1 
    \end{aligned}
  }
  \infer1[\EquationReasoning{}: expand sum]{
    \begin{aligned}
        i = \TransitionVar_5, l_A = \sum_{k = 1}^{k = 5} \TransitionVar_k \land n = \TransitionVar_4 + \TransitionVar_1 \land n \geq i \land l_A - i - n > 0
    \land \Connected(A', \Filter)
    \end{aligned}
  }
  \infer1[\EquationReasoning{}: expand sum]{
    \Connected(A', \Filter) \land
    \left[l_A, i, n\right] =
    \sum\limits_{\Transition \in \Transitions_{A'} } \Map(\Transition) \cdot \Filter(\Transition)
  }
  \infer1[\Expand{}]{\Image{}_{A', \Map}(\Filter, \left[l_A, i, n\right]) \land n \geq i \land l_A - i - n > 0}
\end{prooftree}
\caption{A derivation tree for an application of
\cref{const:s1-in-c-dd,const:s1-substring,const:something-before-and-after,const:more-inside-than-before}
from \cref{ex:string-constraints}, corresponding to handling the Parikh
automaton $A'$ of \cref{fig:aut_a}. Note that the flow equations of
\cref{eq:single-filter} are left out of the tree in order to keep it
manageable.}\label{fig:derivation:single}
\end{figure}

### 2023-06-24 example.tex
An acceptable (but hardly great) example
This also adds propagation to 2.1, which I had forgotten.

Punching out from today after 5h 30 min.

\subsubsection{Propagation of connectivity constraints}\label{sec:intuition:prop}
At this point we can conclude that none of the loops of $A'$ can become
disconnected since they are at the initial and final states, both of which are
reachable for any future assignments of the transition variables. This means
that all valid counter assignments of $A'$ are described by
\cref{eq:counter-sums} and the flow equations alone without the need for
expensive connectivity constraints.

\subsubsection{Case Splitting}\label{sec:intuition:split}
After concluding this, we have a choice of two paths through automaton $B$; the
upper through state $A$ or the lower through state $B$. We case split on a
problem-aware basis by selecting a transition variable that would disconnect
some strongly connected component of automaton $B$, in this case the transition
form $I_B$ to state $B$. We divide it into the cases used ($> 0$) or unused ($ =
0$) since those are the threshold values of the calculus, preferring the latter
option on the first-fail principle.

>>>>

\subsubsection{Subsumption of connectivity constraints}\label{sec:intuition:subsume}
At this point we can conclude that none of the loops of $A'$ can become
disconnected since they are at the initial and final states, both of which are
reachable for any future assignments of the transition variables. This means
that all valid counter assignments of $A'$ are described by
\cref{eq:counter-sums} and the flow equations alone without the need for
expensive connectivity constraints.

\subsubsection{Case Splitting and propagation of connectivity constraints}\label{sec:intuition:split}
After concluding this, we have a choice of two paths through automaton $B$; the
upper through state $A$ or the lower through state $B$. We case split on a
problem-aware basis by selecting a transition variable that would disconnect
some strongly connected component of automaton $B$, in this case the transition
form $I_B$ to state $B$. We divide it into the cases used ($> 0$) or unused ($ =
0$) since those are the threshold values of the calculus, preferring the latter
option on the first-fail principle.

Having performed the split, we can conclude that the path from $I_B$ to state
$B$ is now unreachable without passing through a zero-times-used transition.
This means that the self-loop from state $B$ is also unusable, and thus must
also be used zero times.
